Composition-climate paper

### 2015-10-27 Composition_Ecology.Rmd
Some editing in the abstract and introduction, but snall changes, mostly changing comma separated references to semicolon separated.
Added more text to the 'regional climate change' section and a few references from Viau, and Ahmed.  Could add more, maybe plumb the PalEON climate paper.

Beefed up the caption for Figure 5.

Knit with the changes.

*Predictions for future species range shifts are predicated largely on correlative models that relate modern species distributions to climate parameters in the modern era.  Given the extent of regional forest change and recent losses to key forest taxa including Hemlock, Elm and Chestnut it is possible that certain taxa have shiften with respect to climate change over the last century.  The additional pressure of climatically biased land use conversion for agriculture along the western and southern border of the upper Midwest means that the shift in climate space may be most pressing in a region of climate space that is critically important for understanding future species responses to climate change.*

@@

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies,@schulte2001original,@schulte2007homogenization,@radeloff2000historical]. To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies], we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

@@

Methods
===============
We pair pre-settlement vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLS data, Goring et al. [-@goring2015] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

Another complication is the structure of the FIA dataset.  While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree,@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations. [needs refs].

@@

The choice of a thirty year normal for the PLS data is potentially problematic.  The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset, @white1983history], and for much of this time period no climatological records exist. Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period. As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.  Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions [@trouet, @mann] tell us _THIS_.

@@

**Figure 5**. *For each of the four variables examined we see...*

@@

The precipitation panel (Figure X) indicates that precipitation across the region has driven change in the species climate niche for this variable.  Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.

>>>>

*Predictions for future species range shifts are largely predicated on correlative models that relate modern species distributions to climate parameters in the modern era.  Given the extent of regional forest change, and recent losses to key forest taxa including Hemlock, Elm and Chestnut, it is possible that certain taxa have shifted their climatic distributions over the last century.  The additional pressure of climatically biased land use conversion for agriculture along the western and southern margins of the upper Midwest means that the shift in climate space may be most pressing in a region of climate space that is critically important for understanding future species responses to climate change.*

@@

In the midwestern United States we see a strong gradient of temperature and precipitation that results in two major ecotones, one from prairie/savanna to closed forest, in an approixmately west to east direction, and one from conifer dominated, sub-boreal forests to decidious forest in a north to south direction.  Goring et al. [@goring2015preset] have shown that land use change has significantly changed the strength and structure of this ecotone, and work across this regions has consistently shown a significant imprint of human land use change on this region [@rhemtulla2009legacies;@schulte2001original;@schulte2007homogenization;@radeloff2000historical]. To date researchers have not examined how these shifts might affect the relationships between climate and the distributions of individual taxa. Given that much of the land use change within the region occurs in the south, where agricultural conversion has largely eliminated open forests [@rhemtulla2009legacies], we might expect to see that species with more southerly distributions would show greater impacts of land use change on their distributions in climate space, assuming no shift in regional climate.  Conversely, species in the north should show little change if we expect that pre-settlement trees show similar regeneration patterns (with respect to simple presence/absence) following widespread logging.

@@

Methods
===============
We pair pre-settlement vegetation data from the Public Land Survey System (PLSS) aggregated to a 64 $km^{2}$ grid [@goring2015preset] with annually resolved climate data for the region.  The PLSS era data is aggregated from survey data collected on a 1 x 1 mi grid across the Upper Midwest during the PLS, from 1830 - 1904, in a time transgressive manner, from the south east to the north west.  At each survey point surveyors noted the closest two to four trees, recorded their distances to the plot center point, diameters and the azimuth to the tree.  Goring et al. [-@goring2015preset] aggregated this data to an 8 x 8 km grid to reduce local-scale variability, resulting in a gridded data product with an average of 61 plots per cell, or approximately 120 trees per cell.  When examining relationships between FIA and PLS data, Goring et al. [-@goring2015preset] report significant relationships within the FIA relating dissimilarity from PLSS era forests and the number of FIA plots within an 8x8km cell, however this relationship accounts for only 3% of the total change in dissimilarity, indicating statistical significance, but little ecological significance on this regional scale.

Another complication is the structure of the FIA dataset.  While the PLS data was collected across the western US on a uniform grid the FIA data is only collected from forested or partially forested landscapes [@woudenberg2010forest].  As such trees may be present in regions without permanant or temporary FIA plots, but no record of them exists within the FIA. This complicates our ability to assign 'absence' within the FIA, however the practice of using FIA data as a surrogate for presence/absence data is widespread [@iverson2013tree;@iverson1998predicting], and, at a broad scale, the presence of individual and isolated trees in unforested agricultural landscapes may be significant for adaptation, but it may also be the case that these trees remain on the landscape through anthropogenic agency, for example, as a result of irrigation or landscaping care within urban or rural locations. [needs refs].

@@

The choice of a thirty year normal for the PLS data is potentially problematic.  The time transgressive nature of the PLS Survey means that some of the PLS data was sampled in the early 1800s, while other come from the early 1900s [@goring2015preset;@white1983history], and for much of this time period no climatological records exist. Regardless, throughout this period climate variability exists (Figure 1d), and trees that were on the landscape from the 1830s to the 1880s are not expected to have been completely extirpated by annual scale climate variability in that same time period. As such the use of a 'pre-settlement' climate normal from 1895-1925 is likely a reasonable choice.  Regional-scale reconstructions from tree-rings and from pollen-based climate reconstructions tell us that there has been a trend of increasing temperatures (approximately 0.3^o^C in the Northern Hemisphere) since the 1800s, with pollen-based reconstructions showing even higher rates of change in the Midwestern United States [@viau2006millennial], with tree rings showing cooling at the 1800s to warming from the 1900s [@ahmed2013continental].

@@

**Figure 5**. *For each of the four variables examined we see differences in the relative contributions of land use and climate in driving patterns of change over the settlement era.  Taxa with negative y values have changes in distributions along the climate gradient that are driven largely by land use, while positive y axis values can be largely attributed to regional climate change.*

@@

The precipitation panel (Figure 5) indicates that precipitation across the region has driven change in the species climate niche for this variable.  Only spruce and poplar show a marginally greater effect of land use change than climate change, wheras taxa such as beech and hemlock show a strong climate signal.
